A Linear Solution of the Linearity Challenge

### 2025-12-25 paper/related.tex
global revision

\citet{CastroFerreiraYoshida20} describe EMTST, a library for the formalisation
of session type systems which includes as case studies the session calculus of
\citet{HondaVasconcelosKubo98} (called ``original system'') and a revised
version of it that is more amenable to be formalised using a locally nameless
representation of channels.

@@

Motivated by the technical difficulties arising from context splits,
\citet{SanoKavanaghPientka23} define a structural version of \CP using an
alternative approach based on \emph{linearity predicates}. The key idea is to
treat typing contexts structurally and to enforce the linear usage of channels
by checking their syntactical occurrence in processes. Interestingly, the
approach relies on the \emph{explicit naming of continuations} so as to
precisely account for the number of times a channel is actually used.

@@

\Cref{tab:comparison} reports a summary of the aforementioned formalisations and
our own. The three leftmost colums list the reference papers (sorted by
publication date), the proof assistant and the formalised calculus.
%
For simplicity, we have grouped the calculi into classes identified by an
acronym. Some acronyms, like \CP~\cite{Wadler14} and
\GV~\cite{GayVasconcelos10,Wadler14} are well known in the literature on session
types. It should be noted that the actual calculus being formalised usually
differs from (typically, is a strict subset of) the one identified by the
acronym, but it retains at least some of its key features. We use \SessionPI to
denote (variants of) the session-based $\pi$-calculus presented by
\citet{HondaVasconcelosKubo98} and \LinearPI to denote (variants of) the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99}, into which binary sessions can be
encoded using explicit continuation channels as described by \citet[extended
version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}. Finally, we use
\SCP for the structural version of \CP introduced by
\citet{SanoKavanaghPientka23} and \Calculus for our calculus, a linear version
of \CP with explicit continuations.

The four middle columns of \Cref{tab:comparison} report the relevant qualitative
aspects of the formalisations, namely the technique used to enforce the linear
usage of channels, the technique used to encode channels, whether the
formalisation is intrinsically or extrinsically typed, and whether the calculus
makes use of continuation channels. Concerning the use of continuations, the
approaches based directly on the linear $\pi$-calculus (into which sessions can
be encoded) are marked with {\YES} and those based on a calculus with native
sessions are marked with {\NO}. The calculus \SCP is marked with {\MAYBE}
because, while not being strictly based on the linear $\pi$-calculus, it makes
use of explicit continuations for defining the linearity predicates that check
the linear usage of channels. Finally, all the approaches based on \GV are also
marked with \MAYBE. In principle, \emph{none} of these calculi makes use of
continuations. However, \GV is designed in such a way that each session
operation acting on a channel $s$ returns the result of the operation (if any)
\emph{along with the same channel $s$}. In this way, the type of $s$ can be
``updated'' to take into account the effect of the operation. As observed by
\citet{Padovani17}, this way of modeling the communication primitives is
virtually indistinguishable from one making use of explicit continuation
channels.

@@

In general, it is difficult to draw firm conclusions on the effectiveness of the
various approaches in addressing the linearity challenge because the
formalisations differ widely for a variety of entangled factors including the
reference calculus, supported features, proof assistant,
process/expression/channel representation, handling of linearity. Also, most
formalisations predate the publication of the linearity
challenge~\cite{CarboneEtAl24}, hence they are not necessarily targeted to
solving that particular problem. However, some of
them~\cite{RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,SanoKavanaghPientka23,ZackonEtAl25}
are explicitly motivated by the complexity arising from the linear usage of
resources/channels and from context splitting in particular.

Based on the available data, we can make the following observations.
%
The use of context splitting, which is very well represented, does not seem to
be a good indicator of the complexity of the formalisation, since it spans the
whole range of sizes. Our formalisation is based on context splitting and it is
the only one that supports all the featurs of \CP, yet it is also the smallest
one.

The largest formalisations~\cite{GotoEtAl16,CastroFerreiraYoshida20} are also
the ones that adopt a locally nameless representation of channels. In these
formalisations channels are represented in two different ways, depending on
whether they are free or bound. This necessarily entails some duplication of
effort as well as some transformation machinery between the two representations
that contribute to the complexity of the formalisation.
%
Other channel representations are not strong complexity indicators. Note that
the adoption of co-de Bruijn syntax implies the use of context splitting, hence
the two aspects are not completely independent.

There is no strong evidence that the intrinsically typed representation of
processes reduces the size of the formalisation. As observed in
\Cref{sec:process-agda}, this choice may help reducing the overall number of
datatypes to be defined, but the definitions are also more involved because they
incorporate invariants and the proofs of typing preservation. In the end, it
seems plausible to speculate that the effort for representing processes, types
and typing rules is not substantially different overall.

Finally, it seems that the use of (explicit) continuations helps taming the
complexity of the formalisation more than anything else. Indeed, the six
smallest formalisations (excluding the deadlock freedom results) -- with an
average size of around 62kb -- are all based on continuations, whether they are
explicit (\LinearPI, \SCP, \LCP) or ``virtual'' (\GV), while the remaining ones
are 263kb on average. This observation is in agreement with the considerations
we made earlier in this paper concerning the difference between \emph{updating}
bindings versus \emph{removing and adding} bindings in a typing context. The
adoption of continuations enables a clean linear management of channels and
avoids the need for strong updates on the typing context.

>>>>

\citet{CastroFerreiraYoshida20} describe EMTST, a library for the formalisation
of session type systems that includes as case studies the session calculus of
\citet{HondaVasconcelosKubo98} (called ``original system'') and a revised
version of it that is more amenable to be formalised using a locally nameless
representation of channels.

@@

Motivated by the technical difficulties arising from context splits,
\citet{SanoKavanaghPientka23} define a structural version of \CP using an
alternative approach based on \emph{linearity predicates}. The key idea is to
treat typing contexts structurally and to enforce the linear usage of channels
by checking their syntactical occurrence in processes. Interestingly, the
approach relies on the \emph{explicit naming of continuations} so as to
precisely account for the number of times a channel is actually used.
\citet{SanoKavanaghPientka23} do not connect their technique with the
continuation-passing encoding of binary sessions, but \emph{a posteriori} the
analogies between the two approaches are evident.

@@

\Cref{tab:comparison} reports a summary of the aforementioned formalisations and
also our own. The three leftmost colums list the reference papers (sorted by
publication date), the proof assistant and the formalised calculus.
%
For simplicity, we have grouped the calculi into classes identified by an
acronym. Some acronyms, like \CP~\cite{Wadler14} and
\GV~\cite{GayVasconcelos10,Wadler14} are well known in the literature on session
types. It should be noted that the actual calculus being formalised usually
differs from (typically, is a strict subset of) the one identified by the
acronym, but it retains at least some of its key features. We use \SessionPI to
denote (variants of) the session-based $\pi$-calculus presented by
\citet{HondaVasconcelosKubo98} and \LinearPI to denote (variants of) the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99}, into which binary sessions can be
encoded using explicit continuation channels as described by \citet[extended
version]{Kobayashi02b} and \citet{DardhaGiachinoSangiorgi17}. Finally, we use
\SCP for the structural version of \CP introduced by
\citet{SanoKavanaghPientka23} and \Calculus for our calculus, a linear version
of \CP with explicit continuations.

The four middle columns of \Cref{tab:comparison} report the relevant qualitative
aspects of the formalisations, namely the technique used to enforce the linear
usage of channels, the technique used to encode channels, whether the
formalisation is intrinsically or extrinsically typed, and whether the calculus
makes use of continuation channels. Concerning the use of continuations, the
approaches based directly on the linear $\pi$-calculus (into which sessions can
be encoded) are marked with {\YES} and those based on a calculus with native
sessions are marked with {\NO}. The calculus \SCP is marked with {\MAYBE}
because, while not being strictly based on the linear $\pi$-calculus, it makes
use of explicit continuations for defining the predicates that check the linear
usage of channels. Finally, all the approaches based on \GV are also marked with
\MAYBE. In principle, \emph{none} of these calculi makes use of continuations.
However, \GV is designed in such a way that each operation acting on a channel
$s$ is a function that returns the result of the operation (if any) \emph{along
with the same channel $s$}. In this way, the type of $s$ can be conveniently
``updated'' to take into account the effect of the operation. As observed by
\citet{Padovani17}, this semantics of the communication primitives is virtually
indistinguishable from one making use of explicit continuation channels.

@@

In general, it is difficult to draw firm conclusions on the effectiveness of the
various approaches in addressing the linearity challenge because the
formalisations differ widely for a variety of entangled factors including the
features of calculus, the used proof assistant, the representation of
processes/expressions/channels, the handling of linearity. Also, most
formalisations predate the publication of the linearity
challenge~\cite{CarboneEtAl24}, hence they are not necessarily targeted to
solving that particular problem. However, some of
them~\cite{RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,SanoKavanaghPientka23,ZackonEtAl25}
are explicitly motivated by the complexity arising from the linear usage of
resources/channels and from context splitting in particular.

Based on the available data, we can make the following observations.
%
The adoption of context splitting, which is very well represented, does not seem
to be a good indicator of the complexity of the formalisation. Indeed, the
formalisations based on context splitting span the whole range of sizes, from
the largest by \citet{GotoEtAl16} (543kb) to our own (21kb, without the proof of
deadlock freedom) which is also the only one supporting all the featurs of \CP.

The two largest formalisations~\cite{GotoEtAl16,CastroFerreiraYoshida20} are the
ones that adopt a locally nameless representation of channels. In these
formalisations channels are represented in two different ways, depending on
whether they are free or bound. This necessarily entails some duplication of
effort as well as some transformation machinery between the two representations
that contribute to the complexity of the formalisation.
%
Other channel representations are not strong complexity indicators. Note that
the adoption of co-de Bruijn syntax implies the use of context splitting, hence
the two aspects are not completely independent.

We found no strong evidence that the intrinsically typed representation of
processes reduces the size of the formalisation. As observed in
\Cref{sec:process-agda}, this choice may help reducing the overall number of
datatypes to be defined, but the definitions are also more involved because they
incorporate invariants and the proofs of typing preservation. In the end, it
seems plausible to speculate that the effort for representing processes, types
and typing rules is not substantially different overall.

Finally, it seems that the use of (explicit) continuations affects the
complexity of the formalisation more than anything else. Indeed, the six
smallest formalisations (excluding the deadlock freedom results) -- with an
average size of around 62kb -- are all based on continuations, whether they are
explicit (\LinearPI, \SCP, \LCP) or ``virtual'' (\GV), while the remaining ones
are 263kb on average. This observation is in agreement with the considerations
we made earlier in this paper concerning the difference between \emph{updating}
bindings versus \emph{removing and adding} bindings in a typing context. The
adoption of continuations enables a clean linear management of channels and
avoids the need for strong updates on the typing context.
